Wrattler: \textnormal{Making notebooks reproducible, live and smart}

### 2018-03-18 tapp/paper.tex
Finishing TaPP draft

\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}

@@

In standard notebook systems, such as Jupyter, the state and execution is handled by a 
\emph{kernel}. The notebook running in a browser sends commands to the kernel to evaluate 
cells selected by the user. As illustrated in Figure~\ref{fig:arch}, Wrattler splits the 
server functionality between the following components:

@@

The Wrattler architecture supports languages that require running a separate language runtime
as a server, but also languages that can be fully evaluated in the browser. To illustrate this
feature, we integrated Wrattler with TheGamma [X], a simple browser-based language for data
exploration.

As an example, we use a data set with broadband speed published by the UK government [X].
The following TheGamma script loads the raw CSV file, groups rows by whether they represent
rural or urban area and then calculates average download speed for both of the kinds of areas:
%
\begin{equation*}
\begin{array}{l}
\ident{web.loadTable}(\str{https://.../bb2014.csv}).\ident{explore}.\\
\quad\qident{group data}.\qident{by Urban/rural}.\\
\qquad\qident{average Download speed (Mbit/s) 24 hrs}  
\end{array}
\end{equation*}
%
Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} [X] 
and the user typically interacts by selecting one of generated members through an auto-complete
(hence the long explicity names are acceptable). For the purpose of this paper, the most 
important aspect is that TheGamma scripts can be parsed, checked and evaluated in the browser.

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\hspace{-6em}\begin{array}{l}
\kvd{procedure}~\ident{bind}(\textit{cache}, \texit{syn})~=\\
\quad \kvd{let}~h = \{ \ident{hash}(\ident{kind}(\textit{syn})) \} \\  
\quad \qquad \cup~ \{\ident{hash}(c) \,|\, c\in\ident{antecedents}(\textit{syn}) \}\\
\quad \kvd{if}~\ident{not}~\ident{contains}(\textit{cache}, h)~\kvd{then}\\
\quad \quad \kvd{let}~n = \emph{fresh node}\\
\quad \quad \ident{value}(n) \leftarrow \ident{Unevaluated} \\
\quad \quad \ident{hash}(n) \leftarrow h\\
\quad \quad \ident{set}(\textit{cache}, h, n) \\
\quad \ident{lookup}(\textit{cache}, h)\\
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{\small{When binding a graph node to a syntactic element, Wrattler first computes
  a set of hashes that uniquely represent the node. This includes hash of the kind of the 
  node (e.g. the source code of a Python node or member name in TheGamma) and hashes
  of all antecedents. If a node with a given hash does not exist in cache, a new node
  is created. We set its hash, indicate that its value has not been evaluated and
  add it to the cache.}}
\label{fig:bind}
\end{figure}

@@

As an example, \emph{datadiff} is an AI assistant that reads two specified data frames from the
data store and suggests a script that transforms the data in the second data frame to match the
format of the first data frame. The AI assistant does not automatically transofrm the data, but
instead, produces a new cell in TheGamma script that can be reviewed by the data analyst. 
For example:
%
\begin{equation*}
\begin{array}{l}
\ident{datadiff(broadband2014, broadband2015)}\\
\quad.\ident{drop\_column}(\str{WT\_national})\\
\quad.\ident{drop\_column}(\str{WT\_ISP})\\
\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
\end{array}  
\end{equation*}
%
The script specifies that two columns should be dropped from the corrupted second data frame and
one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
\strf{Urban} and \strf{Rural}).

\section{Results}
\label{sec:results}

\paragraph{Properties.}
The dependency graph enables several features that are difficult for most notebook systems:
%
\begin{itemize}
\item[--] When code changes, Wrattler only needs to recompute a small part of the graph.
  This makes it possible to provide live previews, especially for simple data analytical DSLs.
\item[--] Refactoring can extract parts of the notebook that
  are needed to compute a specified output.
\vspace{0.25em}
\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
\vspace{0.25em}
\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
  or code cleanup.
\end{itemize}

Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
Once it is interactively constructed, it can be translated to multiple supported languages
such as R, Python or JavaScript.

@@

dataflow notebooks

\newpage
~
\newpage

Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice 
based on statistical analysis of the data and discover hidden structures or corner cases, but only
a human can decide what those mean.
A data science environment of the future thus needs to support an interaction model that 
keeps the human in the loop, allows an efficient human-computer interaction 
and can be extended with new AI asistants that provide advice about data.
In this report, we present Wrattler, a notebook system that is:

\paragraph{Interactive.}
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state.
A data store tracks the provenance and semantics of data, supports versioning and keeps 
the history, making the data analyses fully reproducible.

\paragraph{Polyglot.}
Multiple languages can be used in a single notebook and share data via the data store.
Analysts can use R and Python, but also interactive languages for data exploration
that run in the browser and provide live previews.

\paragraph{Smart.}
Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
help with data cleaning and joining, but also help data exploration by finding typical and atypical 
data points and automatically visualizing data.

\section{Wrattler components}

Wrattler consists of a notebook user interface running in the web browser, which communicates with 
a number of server-side components including language runtimes, the data store and AI assistants. 
In this section, we discuss the components in more detail, starting with the dependency graph
that is maintained on the client-side, by the web browser.

\subsection{Dependency graph}

The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
depend on a data frame (hexagons) exported by an earlier cell. 
When code in a cell changes, Wrattler updates the dependency graph, keeping existing
nodes for entities that were unaffected by the change.

The dependency graph enables several features that are difficult for most notebook systems:
%
\begin{itemize}
\item[--] When code changes, Wrattler only needs to recompute small part of the graph.
  This makes it possible to provide live previews, especially for simple data analytical DSLs.
\vspace{-0.85em}
\item[--] Refactoring can extract parts of the notebook that
  are needed to compute a specified output.
\vspace{0.25em}
\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
\vspace{0.25em}
\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
  or code cleanup.
\end{itemize}

\subsection{AI assistants}

>>>>

\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{flushend}

\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}

@@

In standard notebook systems, such as Jupyter \cite{jupyter}, the state and execution is handled by a 
\emph{kernel}. The notebook running in a browser sends commands to the kernel to evaluate 
cells selected by the user. As illustrated in Figure~\ref{fig:arch}, Wrattler splits the 
server functionality between the following components:

@@

The Wrattler architecture supports languages that require running a separate language runtime
as a server, but also languages that can be fully evaluated in the browser. To illustrate this
feature, we integrated Wrattler with TheGamma \cite{thegamma}, a simple browser-based language for data
exploration.

As an example, we use a data set with broadband speed published by the UK government.
The following TheGamma script loads the raw CSV file, groups rows by whether they represent
rural or urban area and then calculates average download speed for both of the kinds of areas:
%
\begin{equation*}
\begin{array}{l}
\ident{web.loadTable}(\str{https://.../bb2014.csv}).\ident{explore}.\\
\quad\qident{group data}.\qident{by Urban/rural}.\\
\qquad\qident{average Download speed (Mbit/s) 24 hrs}  
\end{array}
\end{equation*}
%
Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} \cite{providers} 
and the user typically interacts by selecting one of generated members through an auto-complete
(hence the long explicity names are acceptable). For the purpose of this paper, the most 
important aspect is that TheGamma scripts can be parsed, checked and evaluated in the browser.

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\hspace{-6em}\begin{array}{l}
\kvd{procedure}~\ident{bind}(\textit{cache}, \textit{syn})~=\\
\quad \kvd{let}~h = \{ \ident{hash}(\ident{kind}(\textit{syn})) \} \\  
\quad \qquad \cup~ \{\ident{hash}(c) \,|\, c\in\ident{antecedents}(\textit{syn}) \}\\
\quad \kvd{if}~\ident{not}~\ident{contains}(\textit{cache}, h)~\kvd{then}\\
\quad \quad \kvd{let}~n = \emph{fresh node}\\
\quad \quad \ident{value}(n) \leftarrow \ident{Unevaluated} \\
\quad \quad \ident{hash}(n) \leftarrow h\\
\quad \quad \ident{set}(\textit{cache}, h, n) \\
\quad \ident{lookup}(\textit{cache}, h)\\
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{\small{When binding a graph node to a syntactic element, Wrattler first computes
  a set of hashes that uniquely represent the node. This includes hash of the kind of the 
  node (e.g. the source code of a Python node or member name in TheGamma) and hashes
  of all antecedents. If a node with a given hash does not exist in cache, a new node
  is created. We set its hash, indicate that its value has not been evaluated and
  add it to the cache.}}
\label{fig:bind}
\end{figure}

@@

As an example, datadiff \cite{datadiff} is an AI assistant that reads two specified data frames from the
data store and suggests a script that transforms the data in the second data frame to match the
format of the first data frame. The AI assistant does not automatically transofrm the data, but
instead, produces a new cell in TheGamma script that can be reviewed by the data analyst. 
For example:
%
\begin{equation*}
\begin{array}{l}
\ident{datadiff(broadband2014, broadband2015)}\\
\quad.\ident{drop\_column}(\str{WT\_national})\\
\quad.\ident{drop\_column}(\str{WT\_ISP})\\
\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
\end{array}  
\end{equation*}
%
The script specifies that two columns should be dropped from the corrupted second data frame and
one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
\strf{Urban} and \strf{Rural}).

\section{Properties of Wrattler}
\label{sec:results}

The Wrattler architecture outlined in Section~\ref{sec:arch} leads to a number of properties that
are difficult to obtain with traditional notebook systems. In this section, we discuss the 
properties and briefly discuss our prototype implementation.

\subsection{Reproducible, live and smart}

The two most important aspects of the Wrattler architecture are that it separates the state from 
the language runtime (using a data store) and that it keeps a dependency graph based on the current
notebook source code (on the client). This enables a number of properties.

\paragraph{Reproducibility.} The evaluation outputs displayed in Wrattler notebook always reflect
the current source code. When code changes, Wrattler updates the dependency graph and hides 
invalidated visualizations. Because the data store caches earlier results, it is always possible
to go back without re-evaluating the whole notebook.

\paragraph{Provenance.} The graph allows us to track data provenance. For code written in TheGamma
script, we can track provenance at the level of individual variables and operations; for R, Python
or JavaScript, we can currently only track provenance at cell level. This also enables implementing
refactoring of notebooks, for example, to extract only code necessary to produce a given visualization.

\paragraph{Interactivity.} Finally, the use of dependency graph makes it possible to update 
displayed results more efficiently, because only values for new nodes in the graph need to be
calculated. The fine-grained structure of the dependency graph for TheGamma script makes it 
possible to often show live previews instantly.

\paragraph{Polyglot.} Sharing the state via data store makes it possible to combine multiple
langauge runtimes, as long as they support sharing data via data frames. In our prototype, this
includes R, JavaScript and TheGamma script, but the extensibility model allows adding further
languages.

\paragraph{Smart and explainable.} The data store makes it possible to build third-party components, 
that analyze data in the notebook and provide automatic suggestion, such as how to merge entities
across multiple data sets. Furthermore, the data store does not allow direct modification of the 
data, so such tools result in code that explains the operation and can be reviewed by the data
analyst.

\subsection{Wrattler prototype}

A prototype implementation of Wrattler is available on GitHub (\url{http://github.com/wrattler}).
The prototype implements language runtimes for R, JavaScript and TheGamma script 
(Section~\ref{sec:comp-gamma}). It builds a dependency graph (Section~\ref{sec:comp-deps})
and uses it to evaluate results of cells. The data store (Section~\ref{sec:comp-data}) 
stores data in Microsoft Azure in JSON format. Support for meta-data annotations and big data
is not yet implemented. The protoype comes with one third-party tool -- the datadiff AI 
assistant mentioned above (Section~\ref{sec:comp-data}). Figure~\ref{fig:proto} shows the 
first cell of a sample Wrattler notebook.

\begin{figure}
\vspace{-0.5em}
\includegraphics[scale=0.15]{screen.png}
\vspace{-0.5em}
\caption{\small{TheGamma script that downloads and aggregates UK government data, running in 
  Wrattler notebook with a live preview.}}
\label{fig:proto}
\end{figure}

@@

The work in this paper directly follows the work on IPython and Jupyter notebook systems
\cite{ipython}\cite{jupyter}. Wrattler shares many properties with those and aims to address some
of their limitations. To address reproducibility, some Jupyter extensions and systems such
as R markdown \cite{rmarkdown} lock cells after evaluation. Dataflow notebooks 
\cite{dataflow} are more sophisticated and attach unique hashes to cell evaluations.
Scientific workflow systems \cite{taverna,kepler} manage evaluation of workflows over a 
dependency graph similarly to Wrattler, but often allow editing it directly via a GUI, rather
than through code in a notebook.

Our method of dependency grpah construction is inspired by Roslyn \cite{roslyn} and extends our
earlier work on TheGamma \cite{thegamma}. It is simiar to methods used in live programming 
languages \cite{live,subtext}, as well as incremental compilation \cite{incremental} and
partial evaluation \cite{partial}. Wrattler adapts those methods to a notebook-like data
science environment.

\section{Summary}

This paper presents early work on Wrattler -- a new notebook system for data science that makes
notebooks reproducible, live and smart. The properties of Wrattler follow from two changes to 
the standard architecture of notebook systems.

First, Wrattler separates the state management from code execution. This allows provenance 
tracking, polyglot notebooks and integration of third-party tools that can work directly with
the data store. Second, Wrattler keeps a dependency graph on the client (web browser) and uses 
it to control evaluation. This guarantees reproducibility and allows faster feedback during
development.
